---
title: 'Major League Baseball Playoff Appearance Likelihood'
author: "Shannon Leiss"
date: "12/7/2021"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(Lahman)
library(patchwork)
library(ggpubr)
```

## Introduction

Many different attempts have been made to optimize an equation for predicting a team's probability of making the playoffs in all sports. 

`Teams` dataset in the `Lahman` package provides team records for the 1871 - 2020 baseball season, along with offensive statistics, defensive statistics, and if the team made the playoffs. 

Using this data for the seasons of 1998 - 2019, offensive statistics will be used to model the likelihood of a team making the playoffs. 

Based off the best models, the offensive statistics needed to have a 50% and 90% chance of making the playoffs will be found.

Using the same offensive data, it will be compared which league - American or National - hits more home runs and which franchises historically hit the most home runs per season.


## Offensive Statistics Being Considered 

The offensive statistics used initial were:

  - Runs Scored
  
  - Hits

  - At Bats

  - Doubles

  - Triples

  - Homeruns 

Addition variable that were calculated for use in a separate model were:

  - Run Average: Runs Scored divided by At Bats

  - Batting Average: Hits divided by At Bats

## Data Used:

```{r, echo=F}
Season_by_Team <- Teams
Season_by_Team <- Season_by_Team%>%subset(yearID >= 1998)
Season_by_Team <- Season_by_Team %>% mutate(Playoffs = ifelse(DivWin == "Y" | WCWin == "Y",1,0),WS_Appearance = ifelse(LgWin == "Y",1,0),WS_Champ = ifelse(WSWin == "Y",1,0) )
Season_Offensive <- Season_by_Team[Season_by_Team$yearID != 2020,c(1,2,4,5,15:21,26,44,49:51)] 
rownames(Season_Offensive) <- NULL
Season_Offensive <- Season_Offensive%>% mutate(RperAB = round(R/AB,4), HperAB = round(H/AB,4))
New <- t(Season_Offensive[,c(5,6,7,8,9,10,17,18,14)])
colnames(New) <- Season_Offensive$franchID
knitr::kable((New[,1:5]))
```


## Initial Logistic Model 

Since the response variable of `Playoff` is a indicator of whether or not a team made the playoffs, logistic regression can be used to model the log-odds of a team making the playoffs. The initial logistic model that was fitted:

$$\begin{aligned}
log(\frac{p_i}{1-p_i}) =  25.306 &+ 0.0178(R_i) - 0.0068(AB_i)- 0.00015(H_i)\\
&- 0.0035(X2B_i) - 0.0112(X3B_i) - 0.003(HR_i)
\end{aligned}$$
Where $p_i$ is the probability that a team made the playoffs.

This model had an AIC of 673.9, a residual deviance of 659.9 on 653 degrees of freedom, with no evidence of over dispersion.

A model with all possible interactions was also fitted and was found to be worse than the initial model with no interactions.

## Reduced Models

From the initial model, only the intercept, Runs Scored, and At Bats were significant. The reduced models that were fit:

$$\begin{aligned}
log(\frac{p_i}{1-p_i}) &=  26.2955 + 0.0162(R_i) - 0.0071(AB_i)\\
log(\frac{p_i}{1-p_i}) &=  -115.292 + 0.1980(R_i) \\
&\quad - 0.01837(AB_i) - 0.00003(R_i*AB_i)
\end{aligned}$$

The AIC value for the model without interactions was 667.51 and the model with interactions had an AIC of 664.75.

## Comparing Initial and Reduced Models 

Using drop in deviance tests with likelihood ratios(no over dispersion), the difference in deviance(T) was compared to a $\chi^2_d$ distribution, where d is the difference in terms between models.

Comparing the initial model to the reduced model without interactions:

  - T = 1.6161 with a p-value = 0.8059
  
  - **Reduced Model** should be used 

Comparing the reduced model with interactions to the model without interactions:

  - T = 4.763 with a p-value = 0.02908
  
  - **Model with Interactions** should be used


## Logistic Model using Mutated Offensive Statistics

Using the created variables of Run Average and Batting Average per team, the following logistic model with interaction was fit:

$$\begin{aligned}
log(\frac{p_i}{1-p_i}) =  -41.92 &+ 330.99(RperAB_i) + 110.21(HperAB_i)\\
&- 922.93(RperAB_i*HperAB_i)
\end{aligned}$$

This model had an AIC of 674.71 and a residual deviance of 666.71 on 656 degrees of freedom.

Reduced Model without interaction:
$$
log(\frac{p_i}{1-p_i}) =  -7.808 + 87.704(RperAB_i) -19.536(HperAB_i)
$$

With an AIC of 675.32 and a residual deviance of 669.32 on 657 degrees of freedom.


## Initial vs Reduced Model

Using drop in deviance test between above models:

   - T = 2.6124, with p-value = 0.106
   
   - **Model without Interaction** should be used


## Optimal Models

Using Offensive statistics:

$$\begin{aligned}
(1) \qquad log(\frac{p_i}{1-p_i}) &=  -115.292 + 0.1980(R_i) \\
&\quad - 0.01837(AB_i) - 0.00003(R_i*AB_i)
\end{aligned}$$

Using Mutated Offensive statistics:

$$
(2) \qquad log(\frac{p_i}{1-p_i}) =  -7.808 + 87.704(RperAB_i) -19.536(HperAB_i)
$$

## Different Link Functions on Models 

Using the two "best" models, different link functions were tried and compared:
$$\begin{aligned}
\mbox{Probit: }& \Phi^{-1}(p_i) = \alpha + \beta_ix_{ij}\\
\mbox{C-Log-Log: }& log(-log(1-p_i)) = \alpha + \beta_ix_{ij}
\end{aligned}$$

Comparing the AIC values for both models using all three links:
```{r,echo=F}
mod1 <- glm(Playoffs ~ R+AB+R*AB,data=Season_Offensive,family=binomial)
mod2 <- glm(Playoffs ~RperAB + HperAB,data=Season_Offensive,family=binomial)
mod1.1 <- glm(Playoffs ~ R+AB+R*AB,data=Season_Offensive,family=binomial(link="probit"))
mod1.1.reduced <- glm(Playoffs ~ R+AB,data=Season_Offensive,family=binomial(link="probit"))
mod2.1 <- glm(Playoffs ~RperAB + HperAB,data=Season_Offensive,family=binomial(link="probit"))
mod2.1.2<- glm(Playoffs ~RperAB + HperAB+RperAB*HperAB,data=Season_Offensive,family=binomial(link="probit"))
mod1.2 <- glm(Playoffs ~ R+AB+R*AB,data=Season_Offensive,family=binomial(link="cloglog"))
mod1.2.red <- glm(Playoffs ~ R+AB,data=Season_Offensive, family = binomial(link="cloglog"))
mod2.2 <- glm(Playoffs ~RperAB + HperAB,data=Season_Offensive,family=binomial(link="cloglog"))
mod2.2.more <- glm(Playoffs ~RperAB + HperAB+RperAB*HperAB,data=Season_Offensive,family=binomial(link="cloglog"))

AIC.mod1 <- data.frame(mod1$aic, mod1.1$aic, mod1.2$aic)
AIC.mod2 <- data.frame(mod2$aic, mod2.1$aic, mod2.2.more$aic)
colnames(AIC.mod1) <- c("Logistic", "Probit", "C-Log-Log")
colnames(AIC.mod2) <- c("Logistic", "Probit", "C-Log-Log")
knitr::kable(rbind(AIC.mod1,AIC.mod2))
```

## Notes on Model Assumptions

All offensive data for a team in a given season were assumed independent.

Also assumed that each teams performance was independent of other teams performance and a given teams performance season to season was also independent.

The response variable of `Playoff` also has dependent observations.


## Model Checking - Residuals

```{r,echo=F}
mod1 <- glm(Playoffs ~ R+AB+R*AB,data=Season_Offensive,family=binomial)
mod2 <- glm(Playoffs ~RperAB + HperAB,data=Season_Offensive,family=binomial)
rd.mod1 <- resid(mod1,"deviance")
rd.mod2 <- resid(mod2,"deviance")
df <- data.frame(Fit = mod1$fitted.values, Residuals =rd.mod1,Season = Season_by_Team[Season_by_Team$yearID != 2020,1])
p1 <- ggplot(df, aes(x=Fit,y=Residuals,col=as.factor(Season)))+
  geom_point(alpha=0.5)+
  ggtitle("Model 1")+
  labs(col="Season")+
  theme(legend.position = "none")
df2 <- data.frame(Fit = mod2$fitted.values, Residuals =rd.mod2,Season = Season_by_Team[Season_by_Team$yearID != 2020,1])
p2 <- ggplot(df2, aes(x=Fit,y=Residuals,col=as.factor(Season)))+
  geom_point(alpha=0.5)+
  ggtitle("Model 2")+
  labs(col="Season")
ggarrange(p1,p2,ncol=1,common.legend = TRUE,legend = "right")
```


## Applying Models to Data

The probability of a team making the playoffs, using both models, is:
$$\begin{aligned}
(1) \qquad p_i &=  \frac{e^{-115.292 + 0.1980(R_i) - 0.01837(AB_i) - 0.00003(R_i*AB_i)}}{1+e^{-115.292 + 0.1980(R_i) - 0.01837(AB_i) - 0.00003(R_i*AB_i)}}\\
(2) \qquad p_i &= \frac{e^{-7.808 + 87.704(RperAB_i) -19.536(HperAB_i)}}{1+e^{-7.808 + 87.704(RperAB_i) -19.536(HperAB_i)}}
\end{aligned}$$

Using these formulas on the 2019 season:
```{r,echo=F}
Offensive_2019 <- Season_by_Team %>% subset(yearID == 2019)%>% mutate(RperAB = R/AB, HperAB = H/AB)
Offensive_2019<- Offensive_2019[,c(15,16,52,53)]
mod1.coefs <- mod1$coefficients
new.data <- data.frame(Inter = rep(1,30), Offensive_2019$R, Offensive_2019$AB, Offensive_2019$R*Offensive_2019$AB)
log_odds <- as.matrix(new.data) %*% t(t(mod1.coefs))
odds <- exp(log_odds)
probs <- odds/(1+odds)
mod2.coefs <- mod2$coefficients
new.data.2 <- data.frame(Inter = rep(1,30), Offensive_2019$RperAB, Offensive_2019$HperAB)
log_odds_2 <- as.matrix(new.data.2) %*%t(t(mod2.coefs))
odds_2 <- exp(log_odds_2)
probs_2 <- odds_2/(1+odds_2)
team_probs_predicted <- data.frame(Team = Season_by_Team[Season_by_Team$yearID == 2019,4], Playoff_Prob_mod_1 = round(probs,4),Playoff_Prob_mod_2 = round(probs_2,4))
colnames(team_probs_predicted) <- c("Team", "Model 1 Probability", "Model 2 Probability")
rownames(team_probs_predicted) <- Season_by_Team[Season_by_Team$yearID == 2019,4]
Team_probs <- t(team_probs_predicted)
knitr::kable(Team_probs[2:3,1:5])
```



## Chance of Making Playoffs 

For a team with an average team batting average(0.2603):

  - A Run Average of .147 achieves 50% chance of going to playoffs
  
  - A Run Average of .172 achieves 90% chance of going to playoffs 

For a team with an average team Run Average(0.1345):

  - A Batting Average of .204 achieves a 50% chance of going to playoffs
  
  - A Batting Average of 0.3166 achieve a 90% chance of going to playoffs 
  
For a team with an average number of At Bats(5542.671):

  - Scoring 806.4 Runs achieves a 50% chance of going to playoffs
  
  - Scoring 937.7 Runs achieves a 90% chance of going to playoffs



## Home Run Rates

Using Poisson regression to see which league, which division, and which team has the highest homerun rates.

All models had evidence of over dispersion - with an over dispersion estimate around 7.5 for all models - and AIC values in the 9000's.

The variable of Ball Park Factor was used an offset in all models.

## American vs National League

Using a quasipoisson regression:
$$
log(\frac{\lambda_i}{BPF_i}) = 0.58768(\mathbf{I}_{AL_i}) +0.51602(\mathbf{I}_{NL_i})
$$

Using observed data it was found:
```{r,ech=F}
AL_rate <- 1.799816
NL_rate <- 1.675343
Rates <- data.frame(AL_rate,NL_rate)
colnames(Rates) <- c("American League", "National League")
knitr::kable(Rates)
```


## Division Comparisons(West vs Central vs East)

Using a quasipoisson regression:
$$\begin{aligned}
log(\frac{\lambda_i}{BPF_i}) = 0.50845(\mathbf{I}_{AC_i}) &+ 0.64126(\mathbf{I}_{AE_i}) + 0.61208(\mathbf{I}_{AW_i})\\
&+ 0.55202(\mathbf{I}_{NC_i}) + 0.50342(\mathbf{I}_{NE_i})\\
&+ 0.48658(\mathbf{I}_{NW_i})
\end{aligned}$$

Using the observed data, it was found:
```{r,echo=F}
Season_by_Team_Updated <- Season_by_Team %>% subset(yearID != 2020)
Season_by_Team_Updated<- Season_by_Team_Updated %>% mutate(lgDiv = ifelse(lgID == "NL" & divID == "W", "NW",
                                                                  ifelse(lgID == "NL" & divID == "C", "NC",
                                                                  ifelse(lgID == "NL" & divID == "E", "NE",
                                                                  ifelse(lgID == "AL" & divID == "W", "AW",
                                                                 ifelse(lgID == "AL" & divID == "C",
                                                                        "AC","AE"))))))
HR_combo_mod_off_over <- glm(HR ~ as.factor(lgDiv)+offset(log(BPF))-1, data=Season_by_Team_Updated, quasipoisson)
rate_AC <- exp(HR_combo_mod_off_over$coefficients[1])
rate_AE <- exp(HR_combo_mod_off_over$coefficients[2])##Has the highest rate
rate_AW <- exp(HR_combo_mod_off_over$coefficients[3])
rate_NC <- exp(HR_combo_mod_off_over$coefficients[4])
rate_NE <- exp(HR_combo_mod_off_over$coefficients[5])
rate_NW <- exp(HR_combo_mod_off_over$coefficients[6])
Rates <- data.frame(Rates = round(c(rate_AC,rate_AE,rate_AW,rate_NC,rate_NE,rate_NW),4))
rownames(Rates) <- c("AL Central", "AL East", "AL West", "NL Central", "NL East", "NL West")
knitr::kable(Rates)

```

## Team Comparisons

Using a quasipoisson model, the homerun rates for each team were modeled. 

Using the observed data it was found the top 5 teams were:
```{r,echo=F}
HR_lgmod_over_off <- glm(HR ~ franchID+offset(log(BPF))-1,data=Season_by_Team_Updated,quasipoisson)
rates <- exp(HR_lgmod_over_off$coefficients)
Team_rates <- data.frame("Rate" = rates,Team = c("ANA","ARI","ATL","BAL","BOS","CHC","CHW","CIN","CLE","COL","DET","FLA","HOU","KCR","LAD","MIL","MIN","NYM","NYY","OAK","PHI","PIT","SDP","SEA","SFG","STL","TBD","TEX","TOR","WSN"))
Team_rates <- Team_rates[order(Team_rates$Rate,decreasing = T),] 
rownames(Team_rates) <- NULL
rates_team <- t(Team_rates)
colnames(rates_team) <- rates_team[2,]
knitr::kable(t(rates_team[1,1:5]))
```




## Conclusions

The likelihood of a team making the playoffs can be modeled using Offensive and mutated Offensive statistics.

For a team with an average batting average, they would need to score 1.72 runs per ever 10 at bats to have a 90% chance of making the playoffs.

The America League has a higher homerun rate than the national league, with the American League East having the highest rate of all divisions. 

An American League East team - New York Yankees - have the highest homerun rate, with 3 of the 5 top teams coming from the American League East and all 5 top teams coming from the American League.






